Supplementary material

### 2022-02-17 analysis/paper/paper.Rmd
Setting up re-dating of shoreline dated sites
All analyses have been re-run using IDW for shoreline displacement

The close association between Scandinavian Stone Age settlements and shifting prehistoric shorelines was becoming apparent by the end of the 19th century [@degeer1896], and was firmly established and first applied as a dating method in Sweden [@hollender1901] and Norway [@brogger1905] at the turn of the century. Shoreline dating has been fundamental to Stone Age research in Norway ever since [e.g. @mikkelsen1975; @berg-hansen2009; @bergsvik2020; @bjerck2008; @damlien2021]. The method is used both independently, and to compliment other sources of temporal data such as typological indicators or radiometric dates. However, given the coarse and fuzzy resolution of established typological frameworks, the vast amount of surveyed sites that only contain generic lithic debitage that could hail from any part of the period, and as the conditions for the preservation of organic material is typically poor in Norway, dating with reference to shoreline displacement is often the only and most accurate method by which one can hope to date the sites. Shoreline dating is consequently fundamental to our understanding of the Norwegian Stone Age. This is both because it is central to the temporal framework on which our understanding of the period is based, but also because the method is only applicable so long as the societies in question have continuously settled on or close to the contemporary shoreline. Consequently, adherence or deviation from this pattern also has major implications for the socio-economic foundations of the societies in question.

Despite its important role for Norwegian Stone Age archaeology, the applicability of dating by reference to shoreline displacement has only been evaluated using relatively coarse methods. The aim of this paper is to provide a systematic and comprehensive review of the degree to which radiocarbon dates correspond with the dates informed by our current knowledge of shoreline displacement in a larger area of south-eastern Norway, using a more refined methodological approach. The goal is to quantify the degree to which the assumption of shore-bound settlement holds through the Stone Age, and in turn have this inform a refined method for shoreline dating. As presented in more detail below, this problem involves the combined evaluation of three major analytical dimensions. One is the questions of when the sites were in use, the second pertains to the reconstruction of the contemporaneous sea-level, and the third follows from the fact that the relation between site and shoreline is inherently spatial. Taking inspiration from studies that have integrated various sources of spatio-temporal uncertainty through Monte Carlo simulation [@bevan2013a; @crema2012; @crema2010; @crema2015; @yubero-gomez2016], a similar approach is adopted here.

@@

Relative sea-level (RSL) can be defined as the mean elevation of the surface of the sea relative to land, or, more formally, the difference in elevation between the geoid and the surface of the Earth as measured from the Earth's centre [@shennan2015]. Variation in this relative distance follow from a range of effects [e.g. @milne2009; @mörner1976]. Of central importance here is eustasy and istostasy. The eustatic sea-level is understood as the sea-level if the water has been evenly distributed across the Earth's surface without adjusting for variation in the rigidity of the Earth, its rotation, or the self-gravitation inherent to the water body itself [@shennan2015]. The eustatic sea-level is mainly impacted by glaciation and de-glaciation, which can bind or release large amounts of water into the oceans, as well as by changes in the Earth's crust that impact the volume of ocean basins. Istostasy, on the other hand, pertains to adjustments in the crust to regain gravitational equilibrium relative to the underlying viscous mantle. This is often the result of glacial istostasy, which follows from glaciation and de-glaciation and corresponding loading and unloading of weight, as well as from erosion of the crust, which causes its weight to be redistributed. These effects thus causes the lithosphere to either subside due to increased weight, or to rebound and lift upwards due to lower weight [@milne2015].

@@

The shore-bound settlement location of prehistoric hunter-fisher-gatherers in Norway is generally believed to follow both from the exploitation of aquatic resources and from movement and communication, which would have been efficient on waterways. The same logic has also been extended to the hinter- and inland regions, where sites are to be located along rivers and lakes [@brogger1905; @glorstad2010, 57--87; but see @mjærum2018; @gundersen2013; @schulke2020]. This is to take a dramatic turn at the transition to the Late Neolithic, around 2400 BCE, with the introduction of the Neolithic proper [@solheim2021; @prescott2020]. The shift to agro-pastoralism and the introduction of the farm led the site locations to be more withdrawn from the shoreline [e.g. @bakka1971; @prescott2012; @ostmo2008, 223]. That is not to say that waterways and aquatic resources were no longer exploited, but rather that these activities would not have been as tightly integrated with settlement and tool-production areas as in preceding periods [@glorstad2011a]. At an earlier stage, at the transition to the Early Neolithic (c. 3900 BCE), pottery is introduced to the sites, and there are some indications of an initial uptake of agriculture at some sites in the Oslo fjord region. However, this appears to be small in scale and is believed to be combined with a continued and predominantly hunter-gatherer life-way, possibly followed by a complete de-Neolithisation in the Middle Neolithic [@nielsen2019; @hinsch1955]. @nielsen2021b has recently argued that the initial uptake of agriculture in the Early Neolithic is combined with a more complex settlement pattern, and that a simple forager/farming dichotomy would underplay the variation present in the Early and Middle Neolithic data [see also @amundsen2006; @solheim2012, 74]. Seen in relation to the question of interest here, the empirical expectation for the above outlined development would thus be a more varied association between sites and the shore-line from around 3900 BCE, followed by a more decisive shift around 2400 BCE with a comprehensive Neolithisation.

[@breivik2014; @solheim2018]

@@

# Data
To get at the relationship between sites and the contemporaneous shoreline, this analysis is dependent on a study area with good control of the trajectory of prehistoric shoreline displacement. While there is displacement data available for other areas of south-eastern Norway [e.g. @hafsten1957; @sørensen1979; @sørensen1999], considerable methodological developments in recent years means that the most well-established displacement curves are from the region stretching from Horten county in the north-east, to Arendal in the south-west. This area has newly compiled displacement curves for Horten [@romundset2021], Larvik [@sørensen2014; @sørensen2014b; @sorenseninpress], Tvedestrand [@romundset2018; @romundset2018b], and Arendal [@romundset2018b].

The employed shoreline displacement curves are all based on the so-called isolation basin method [@romundset2011]. This involves extracting cores from a series of basins situated on bedrock at different elevations beneath the marine limit, and dating the transition from marine to lacustrine sediments in the cores. Each curve is thus construed from a series of cored basins located at different elevations, each representing a high precision sea-level index point. Furthermore, to minimise the impact of variable uplift rates, the basins are located in a as constrained area of the landscape as possible. Following from the morphology of the retreating ice sheet, the uplift is more severe towards the north-east, meaning that this needs to be adjusted for in the case that any basins are located a significant distance from a common isobase perpendicular to this gradient. The confidence bands of the displacement curves and their trajectory are quite complex constructs, and are the integrated result of both expert knowledge and more objectively quantifiable parameters. The reason for this is in part that the curves do not only contain uncertainty as related to radiometric dates, which are well defined, but also hold potential error as related to the interpretation and analysis of sediment cores, the nature and condition of the basin outlets and the adjustment to a common isobase, to name but a few [e.g. @romundset2019; @romundset2011]. For more details and evaluations done for the compilation of each curve, the reader is therefore referred to the individual publications.

@@

Overall, as is indicated by the measures for central tendency and the almost solid line along the 0 mark on the y-axes, the simulations show that most of the sites have been situated close to the shoreline when they were in use (Figure \@ref(fig:results)). When including all dates, also those that were believed to be erroneous in the original reports, some of the sites are situated considerable distances from the shoreline. However, if one accepts the interpretation in the reports that these dates do not date the main occupation of the sites, as indicated by the artefact inventories, the second row of Figure \@ref(fig:results) gives considerable support to the notion that the sites were in use when they were situated on or close to the contemporaneous shoreline. The distances for some of the earliest sites appears somewhat high, but this can likely be explained as the result of the steepness of the displacement curves for the earliest part of the Holocene (Figure \@ref(fig:map-iso)B), which leads the uncertainty in the ^14^C-dates to give a wider possible elevation range for the sea-level. Another immediately striking result is the apparent deviation from the shoreline towards the end of the Stone Age. From around 2500 BCE several sites are situated a considerable distance from the shoreline, and while a couple remain horizontally and topographically close, most appear to be elevated a considerable distance from the sea-level, as indicated on the plot for vertical distance. This corresponds well with the literature and the suggested impact of the Neolithisation process.

@@

Accepting that shoreline dating appears to loose utility around the transition to the Late Neolithic, as indicated by the clear deviation in site location from the shoreline after this, the results for the second row of Figure \@ref(fig:results) is given again in Figure \@ref(fig:results2), excluding all simulation results younger than 2500 BCE. Furthermore, all negative values have here been set to zero, under the assumption that these result from uncertainty or errors in the data, and not actual site locations. The resulting best point estimate for the vertical distance between site and shoreline for the pre-Late Neolithic is given by the median at `r round(median(distances2$vertdist))`m, following from the skewed distribution. 95% of the vertical distance values fall within the range `r round(quantile(distances2$vertdist, 0))`--`r round(quantile(distances2$vertdist, 0.95))`m. That is, for 95% of the cases, the shoreline was situated on or down to `r round(quantile(distances2$vertdist, 0.95))`m below the site location. This is constrained slightly when only considering the Mesolithic dates, with a median offset of `r round(median(distances3$vertdist))`m and with 95% of distance measures within the range `r round(quantile(distances3$vertdist, 0))`--`r round(quantile(distances3$vertdist, 0.95))`m. An exponential function has been fit to the distributions for vertical distance. While it is clear that this does not perfectly match the data, it would seem reasonable from a theoretical perspective to assume that a process of exponential decay explains this relationship. Furthermore, the accumulation of distances-values on the 0m mark likely follow from forcing negative values to zero, from the resolution of the spatial data, and from defining intersecting sea- and site polygon as having a distance of zero. Using the resulting exponential decay function, the sites treated here have been given a shoreline date in Figure \@ref(fig:shoredate).

>>>>

The close association between Scandinavian Stone Age settlements and shifting prehistoric shorelines was becoming apparent by the end of the 19th century [@degeer1896], and was firmly established and first applied as a dating method in Sweden [@hollender1901] and Norway [@brogger1905] at the turn of the century. Shoreline dating has been fundamental to Stone Age research in Norway ever since [e.g. @berg-hansen2009; @bjerck2008; @breivik2014; @nummedal1923; @mikkelsen1975; @solheim2018]. The method is used both independently, and to compliment other sources of temporal data such as typological indicators or radiometric dates. However, given the coarse and fuzzy resolution of established typological frameworks, the vast amount of surveyed sites that only contain generic lithic debitage that could hail from any part of the period, and as the conditions for the preservation of organic material is typically poor in Norway, dating with reference to shoreline displacement is often the only and most accurate method by which one can hope to date the sites. Shoreline dating is consequently fundamental to our understanding of the Norwegian Stone Age. This is both because it is central to the temporal framework on which our understanding of the period is based, but also because the method is only applicable so long as the societies in question have continuously settled on or close to the contemporary shoreline. Consequently, adherence or deviation from this pattern also has major implications for the socio-economic foundations of the societies in question.

Despite its important role for Norwegian Stone Age archaeology, the applicability of dating by reference to shoreline displacement has only been evaluated using relatively coarse methods. The aim of this paper is to provide a systematic and comprehensive review of the degree to which radiocarbon dates correspond with the dates informed by our current knowledge of shoreline displacement in a larger area of south-eastern Norway, using a more refined methodological approach. The goal is to quantify the degree to which the assumption of shore-bound settlement holds through the Stone Age, and in turn have this inform a refined method for shoreline dating. As presented in more detail below, this problem involves the combined evaluation of three major analytical dimensions. One is the questions of when the sites were in use, the second pertains to the reconstruction of the contemporaneous sea-level, and the third follows from the fact that the relation between site and shoreline is inherently spatial. Taking inspiration from studies that have integrated various sources of spatio-temporal uncertainty through Monte Carlo simulation [e.g. @bevan2013a; @crema2012; @crema2010; @crema2015; @yubero-gomez2016], a similar approach is adopted here.

@@

Relative sea-level (RSL) can be defined as the mean elevation of the surface of the sea relative to land, or, more formally, the difference in elevation between the geoid and the surface of the Earth as measured from the Earth's centre [@shennan2015]. Variation in this relative distance follow from a range of effects [e.g. @milne2009]. Of central importance here is eustasy and istostasy. The eustatic sea-level is understood as the sea-level if the water has been evenly distributed across the Earth's surface without adjusting for variation in the rigidity of the Earth, its rotation, or the self-gravitation inherent to the water body itself [@shennan2015]. The eustatic sea-level is mainly impacted by glaciation and de-glaciation, which can bind or release large amounts of water into the oceans [@mörner1976]. Istostasy, on the other hand, pertains to adjustments in the crust to regain gravitational equilibrium relative to the underlying viscous mantle. This is often the result of glacial istostasy, which follows from glaciation and de-glaciation and corresponding loading and unloading of weight, as well as from erosion of the crust, which causes its weight to be redistributed. These effects thus causes the lithosphere to either subside due to increased weight, or to rebound and lift upwards due to lower weight [@milne2015].

@@

The shore-bound settlement location of prehistoric hunter-fisher-gatherers in Norway is generally believed to follow both from the exploitation of aquatic resources and from movement and communication, which would have been efficient on waterways. The same logic has also been extended to the hinter- and inland regions, where sites are to be located along rivers and lakes [@brogger1905; @glorstad2010, 57--87, but see @mjærum2018; @gundersen2013; @schulke2020]. This is to take a dramatic turn at the transition to the Late Neolithic, around 2400 BCE, with the introduction of the Neolithic proper [@solheim2021; @prescott2020]. The shift to agro-pastoralism and the introduction of the farm led the site locations to be more withdrawn from the shoreline [e.g. @bakka1971; @prescott2012; @ostmo2008, 223]. That is not to say that waterways and aquatic resources were no longer exploited, but rather that these activities would not have been as tightly integrated with settlement and tool-production areas as in preceding periods [@glorstad2011a]. At an earlier stage, at the transition to the Early Neolithic (c. 3900 BCE), pottery is introduced to the sites, and there are some indications of an initial uptake of agriculture at some sites in the Oslo fjord region. However, this appears to be small in scale and is believed to be combined with a continued and predominantly hunter-gatherer life-way, possibly followed by a complete de-Neolithisation in the Middle Neolithic [@nielsen2019; @hinsch1955]. @nielsen2021b has recently argued that the initial uptake of agriculture in the Early Neolithic is combined with a more complex settlement pattern, and that a simple forager/farming dichotomy would underplay the variation present in the Early and Middle Neolithic data [see also @amundsen2006; @solheim2012, 74]. Seen in relation to the question of interest here, the empirical expectation for the above outlined development would thus be a more varied association between sites and the shore-line from around 3900 BCE, followed by a more decisive shift around 2400 BCE with a comprehensive Neolithisation.

@@

# Data
To get at the relationship between sites and the contemporaneous shoreline, this analysis was dependent on a study area with good control of the trajectory of prehistoric shoreline displacement. While there is displacement data available for other areas of south-eastern Norway [e.g. @hafsten1957; @sørensen1979; @sørensen1999], considerable methodological developments in recent years means that the most well-established displacement curves are from the region stretching from Horten county in the north-east, to Arendal in the south-west. This area has newly compiled displacement curves for Horten [@romundset2021], Larvik [@sørensen2014; @sørensen2014b; @sorenseninpress], Tvedestrand [@romundset2018; @romundset2018b], and Arendal [@romundset2018b].

The employed shoreline displacement curves are all based on the so-called isolation basin method [@romundset2011]. This involves extracting cores from a series of basins situated on bedrock at different elevations beneath the marine limit, and dating the transition from marine to lacustrine sediments in the cores. Each curve is thus construed from a series of cored basins located at different elevations, each representing a high precision sea-level index point. Furthermore, to minimise the impact of variable uplift rates, the basins are located in a as constrained area of the landscape as possible. Following from the morphology of the retreating ice sheet, the uplift is more severe towards the north-east, meaning that this needs to be adjusted for in the case that any basins are located any significant distance from a common isobase perpendicular to this gradient. The confidence bands of the displacement curves and their trajectory are quite complex constructs, and are the integrated result of both expert knowledge and more objectively quantifiable parameters. The reason for this is in part that the curves do not only contain uncertainty as related to radiometric dates, which are well defined, but also hold potential error as related to the interpretation and analysis of sediment cores, the nature and condition of the basin outlets and the adjustment to a common isobase, to name but a few [e.g. @romundset2019; @romundset2011]. For more details and evaluations done for the compilation of each curve, the reader is therefore referred to the individual publications.

@@

Overall, as is indicated by the measures for central tendency and the almost solid line along the 0 mark on the y-axes, the simulations show that most of the sites have been situated close to the shoreline when they were in use (Figure \@ref(fig:results)). Some of the sites are situated considerable distances from the shoreline when all dates are included, also those that were believed to be erroneous in the original reports, but if one accepts the interpretation in the reports that these dates do not date the main occupation of the sites, as indicated by the artefact inventories, the second row of Figure \@ref(fig:results) gives considerable support to the notion that the sites were in use when they were situated on or close to the contemporaneous shoreline. The distances for some of the earliest sites appears somewhat high, but this can likely be explained as the result of the steepness of the displacement curves for the earliest part of the Holocene (Figure \@ref(fig:map-iso)B), which leads the uncertainty in the ^14^C-dates to give a wider possible elevation range for the sea-level. Corresponding with the literature, another immediately striking result is the apparent deviation from the shoreline towards the end of the Stone Age. From around 2500 BCE several sites are situated a considerable distance from the shoreline, and while a couple remain horizontally and topographically close, most appear to be elevated a considerable distance from the sea-level, as indicated on the plot for vertical distance.

@@

Accepting that shoreline dating appears to loose utility around the transition to the Late Neolithic, as indicated by the clear deviation in site location from the shoreline after this, the results for the second row of Figure \@ref(fig:results) is given again in Figure \@ref(fig:results2), excluding all simulation results younger than 2500 BCE. Furthermore, all negative values have here been set to zero, under the assumption that these result from uncertainty or errors in the data, and not actual site locations. The resulting best point estimate for the vertical distance between site and shoreline for the pre-Late Neolithic is given by the median at `r round(median(distances2$vertdist))`m. 95% of the vertical distance values fall within the range `r round(quantile(distances2$vertdist, 0))`--`r round(quantile(distances2$vertdist, 0.95))`m. That is, for 95% of the cases, the shoreline was situated on or down to `r round(quantile(distances2$vertdist, 0.95))`m below the site location. This is constrained slightly when only considering the Mesolithic dates, with a median offset of `r round(median(distances3$vertdist))`m and with 95% of distance measures within the range `r round(quantile(distances3$vertdist, 0))`--`r round(quantile(distances3$vertdist, 0.95))`m. While the median distances for horizontal and topographic is only m, it also evident that the range of these values is substantially wider than that for vertical distance, illustrating the point that minor variations in the latter can have substantial consequences for these distance measures, depending on the topography.

An exponential function has been fit to the distributions for vertical distance, estimated by means of maximum likelihood. While it is clear that this does not perfectly match the data, it would seem reasonable from a theoretical perspective to assume that a process of exponential decay explains this relationship. Furthermore, the accumulation of distances-values on the 0m mark likely follow from forcing negative values to zero, from the resolution of the spatial data, and from defining intersecting sea- and site polygon as having a distance of zero.  Having derived a exponential decay function for describing the vertical distance between site and shoreline, this can be combined with the displacement data to provide a method for shoreline dating that takes this into account.

Where x is, asdf. This formula is used to shoreline date the same sites from where this relationship was derived in Figure \@ref(fig:shoredate). The Løvås sites and Gunnarsrød 5 deviate expectedly from the shoreline date. Using the method for evaluating degree of contemporaneity from @parnell2008, given in figure , indicates that the shoreline date is conteporenous date with the radiocarbon dates phase for .

Using the resulting exponential decay function, the sites treated here have been given a shoreline date

### 2022-02-17 analysis/supplementary-materials/supplementary.Rmd
All analyses have been re-run using IDW for shoreline displacement

Sites with ^14^C-dates to the Stone Age but for which no spatial data was readily available are Frebergsvik A--C [@mikkelsen1975, although these do not typologically match], Nedre Holtan, Rugtvedt [@odgaard1994], Tangen [@glorstad2005], and Torsrød [@østmo1975]. These sites are therefore not included in the analysis. Furthermore, @nielsen2021 has recently interpreted features from the otherwise younger sites Bratsberg and Larønningen as possibly related to farming activity in the Early Neolithic. However, given the limited number of features, the lack of any artefacts directly related to these, and the therefore somewhat speculative nature of this suggestion, the sites are kept out of the analysis here.

>>>>

Sites with ^14^C-dates to the Stone Age but for which no spatial data was readily available are Frebergsvik A--C [@mikkelsen1975, although these do not typologically match], Nedre Holtan [@sjursjeike], Rugtvedt [@odgaard1994], Tangen [@glorstad2005], and Torsrød [@østmo1975]. These sites are therefore not included in the analysis. Furthermore, @nielsen2021 has recently interpreted features from the otherwise younger sites Bratsberg and Larønningen as possibly related to farming activity in the Early Neolithic. However, given the limited number of features, the lack of any artefacts directly related to these, and the therefore somewhat speculative nature of this suggestion, the sites are kept out of the analysis here.
